Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-08-14 ms/review-responses.tex

%%%%%%
%%
%%  Don't reorder the reviewer points; that'll mess up the automatic referencing!
%%
%%%%%

\begin{minipage}[b]{2.5in}
  Resubmission Cover Letter \\
    {\it Genetics}
\end{minipage}
\hfill
\begin{minipage}[b]{2.5in}
    C.~J.~Battey, \\
    Peter Ralph, \\
    \emph{and} Andrew Kern \\
  \today
\end{minipage}

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are writing to submit a revised version of our manuscript, 
``Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data''.

\vspace{5em}

\noindent \hspace{4em}
\begin{minipage}{3in}
\noindent
{\bf Sincerely,}

\vskip 2em

{\bf 
C.~J.~Battey, Peter Ralph, and Andrew Kern
}\\
\end{minipage}

\vskip 4em

\pagebreak
\setcounter{page}{1}

%%%%%%%%%%%%%%
\reviewersection{AE}

\begin{quote}
    The manuscript admirably explores a lot of consequences of
    isolation-by-distance in the context of a novel model that is easily amenable
    to forward simulation; however, given that this model may be used in a lot of
    future studies based on the precedent set here, there is some concern about the
    model and its support. Reviewers 2 and 3 highlight this in particular (it
    underlies the main 2 points of reviewer 2's review, and the core of Reviewer
    3's comment), and I agree. Whatever can be done to strengthen the standing of
    this model, and/or connect it to more thoroughly studied models, will be
    helpful for the manuscript. The concern would be that there are peculiarities
    of this model that do not generalize well. A new supplemental section or opener
    to the results section establishing the model more thoroughly would make the
    strongest response.
\end{quote}

\plr{I would generally cut down the quoted bits like the above to only what's essential,
     but haven't done that yet.}

 \plr{IMPORTANT: don't reorder or delete "points" below - it messes up the automatic numbering!}

\begin{point}{Line 35:}
    Also cite Wilkins and Wakeley, Genetics 2002; Wilkins 2004
\end{point}

\reply{
}

\begin{point}{\revref}
    "Such models have been used extensively in ecological modeling but rarely in population genetics " Detailing these previous uses via citations and elaboration may help alleviate the major concern about the provenenance of this model and its unique behaviors (see general comments above and R2 and R3 comments).
\end{point}

\reply{
}

\begin{point}{\revref}
    Please describe computation time needed per replicate
\end{point}

\reply{
}

\begin{point}{\revref}
    I read the acknowledgement to the Hearth and Creative Sky Brewing with a sense of familiarity in feeling of gratitude to my own favorite cafes and breweries, but I it's not a great precedent for Acknowledgements to be filled this way. Please cut.
\end{point}

\reply{
    Good point; we have done this.
}

\begin{point}{Figure 4:}
    Show random-mating expectation
\end{point}

\reply{
}

\begin{point}{Figure 3A, S2:}
    Perhaps more revealing to show on log-log scale?
\end{point}

\reply{
}

\begin{point}{Figure S3:}
    Caption seems to be missing detail 
\end{point}

\reply{
}

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
    This study explores biases arising in population-based inference when 1)
    real population samples are coming from spatial habitat with various degree
    of structuring while inference is made assuming random mating population;
    2) imperfect sampling in practice that fails to represent full diversity
    across entire population habitat; 3) phenotypes that vary across geography
    and create spurious associations with genotypes. While earlier studies
    explored the effect of strong structure on population genetic inference and
    GWAS, this work focuses on less extreme scenarios of structuring that
    arises in populations evolving in continuous habitat. By using
    non-Wright-Fisher model, authors simulated chromosome-scale samples from
    populations that evolved in continuous space, and that can model
    environmental factors to create phenotypes varying over space. As a result,
    this study identified spatial structuring scenarios (small neighbourhood
    size ~ 10-100) that coupled with imperfect sampling strategies lead to a
    biased inference of widely used population genetic statistics (altogether
    18 statistics) such as pi (average pairwise sequence differences),
    heterozygosity (and inbreeding coefficient), and IBS tract sharing.
    Accordingly, inference of the effective population size history was also
    strongly affected under these parameter ranges. Finally, the authors use
    their spatial modelling to demonstrate that typical GWAS with PC-based
    correction cannot entirely remove spurious signals of genotype-phenotype
    associations arising from purely environmental factors.  Overall, the
    authors explore an important but often neglected source of bias that can
    affect inference in many population-based studies (in medical genetics,
    evolutionary biology and ecology). This study can be of interest to a
    broader audience of readership, and I have only minor comments to improve
    clarity and increase accessibility for readers:
\end{quote}

\begin{point}{}
    When neighbourhood size is small (10-100), the mean number of IBS tracts > 2bp (nIBS as in Table S1) is elevated similar to Wright's inbreeding coefficient, but mean of the distribution of pairwise IBS (mean(IBS)) is decreased. What could be the source of this discrepancy? How exactly mean(IBS) was calculated?
\end{point}

\reply{
}

\begin{point}{}
    The authors use K to denote both carrying capacity \llname{ll:K_defn}
    and population density \llname{ll:K_again}. 
    It might be better to use a different notation for these quantities since carrying capacity is fixed while density is an emergent quantity in the non-Wright-Fisher model. Use of K to denote carrying capacity and density is a bit confusing. For example, on \llname{ll:K_more} it is said that ' the "population density" (K) and "mean lifetime" (L) parameters were the same in all simulations'. Here K seems to indicate carrying capacity rather than density? The latter is an emergent quantity and varies across simulation runs?
\end{point}

\reply{
    We agree that this distinction is worth emphasizing!
    We've adjusted our language to hopefully remind the reader that
    $K$ is a parameter that controls population density, rather than being equal to it,
    at \llname{ll:K_again} and \revref{} and \llname{ll:K_more}.
}

\begin{point}{}
    Concerning the non-Wright-Fisher model used, it would be helpful to emphasize that some of the parameters are emergent in contrast to Wright-Fisher model. For example, on Page 11, lines 306-308, the author's goal was to look at census size variation and variation in other quantities. This would be better understood if to emphasize that these parameters are emergent properties in the non-Wright-Fisher model used.
\end{point}

\reply{
}

\begin{point}{}
    Page 9, line 242, Perhaps 'Demographic Inference' might better reflect the content of this section.
\end{point}

\reply{
}

\begin{point}{\revref}
    This sentence with 'Gaussian noise with mean zero and standard deviation 10' is confusing since it was mentioned earlier that the modelled phenotype must vary as human height across Europe, and human height varies 2 standard deviations. Only after reading the whole paragraph it becomes clear that 'standard deviation 10' here referes to unit of height. Please consider rephrasing this sentence.
\end{point}

\reply{
}

\begin{point}{\revref}
    In the sentence, 'We also examined p values for systemic inflation' I think the authors meant 'systematic inflation'.
\end{point}

\reply{
    Whoops; thanks. Fixed.
}

\begin{point}{}
    Page 11, Please correct the legend in Figure 2: must be 'spatial model' and 'random mating' model.
\end{point}

\reply{
}

\begin{point}{}
    Optional: a dashed line in Figure 2 that shows the total carrying capacity of 50*50*5=12500 would be helpful.
\end{point}

\reply{
}

\begin{point}{}
    Page 13, line 349, The phrase 'affect summaries of variation' is better to replace with 'summaries of genetic variation'.
\end{point}

\reply{
}

\begin{point}{}
    Please add or correct references to supplementary figures: For example, Figure S2 was probably meant to accompany Figure 3A, while Figure S1 Figure 3B, but references in the text are absent. In fact, the first reference is made to Figure S3 on page 15.
\end{point}

\reply{
}

\begin{point}{}
    There are also several typos and errors in the text. For example, on Page 12, lines 309; Page 27, line 655.
\end{point}

\reply{
}

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
    Battey et al. use spatially explicit population genetic simulations to
    analyze the effects of spatial structure on (i) the estimation of key
    population genetic parameters, in turn used to (ii) make inferences about
    population history, and on (iii) confounding in genome-wide association
    studies (GWAS). I Liked the paper a lot. It's interesting, well-written and
    addresses an important question - the effect of spatial population
    structure on population genetic statistics and inference-and I enjoyed
    reading it. The most positive aspects were:

    \begin{enumerate}
        \item It nice to actually see spatially explicit simulations and I'm happy that forward simulation is now fast enough that you can do this sort of thing.
        \item The paper is very clear and well-written, easy to understand the motivation and most of the details. That's not always the case for this sort of paper.
        \item I felt that the section about the effect on GWAS was the most interesting and novel part of the paper and gave me some intuition that I hadn't had before.
    \end{enumerate}

    I don't have any major criticisms. There were a few aspects that I thought might warrant some additional discussion, and a few specific questions below. The general questions I had after reading it were:
\end{quote}

\begin{point}{}
    To what extent are any of the results dependent on the exact method of simulation. There are a number of choices about the exact details of the simulations (e.g. the way the overlapping generations are handled, the edge effects and, particularly, the form of Equation 1 - see below). It's not so much that these are non-standard (since I don't think there is a standard) and they all sort of make sense heuristically, and I was left wondering whether these sorts of choices actually make a difference. Do the authors have some thoughts/intuition/results about that? Given that the results in Fig. 3 seem quite consistent with expectations, I suspect that on some level it doesn't make much difference but then there are intermediate results like Fig. 2 which seem a bit counter-intuitive and I wonder if those aspects depend on the simulation scheme.
\end{point}

\reply{
}

\begin{point}{}
    Related to the first point, to what extent are the results qualitatively different to those that would be obtained in a stepping-stone model? My interpretation is that they are actually very similar, but I didn't see whether that was explicitly discussed. In some sense, it's still easier to do large simulations in a stepping-stone model so it would be nice to be reassured that that's still ok.
\end{point}

\reply{
}

\begin{point}{}
    The source of equation (1) is not obvious to me. I sort of see how it makes sense, but a little but more intuition or a brief derivation or an illuminating either in the main text or the supplement, would be helpful.
\end{point}

\reply{
}

\begin{point}{}
    The authors use a scaling factor in equation (2) to counteract the increase in fitness of individuals at the edges. Can they provide a figure showing that this is the case. What does "roughly" mean on line 164. Perhaps a heatmap of the fitness of individuals across the grid with and without the scaling factor?
\end{point}

\reply{
}

\begin{point}{}
    It would be helpful provide the figure showing that generating mutations during the forward simulations in SLIM is equivalent to applying mutations using msprime on pre-generated trees (line 185)? It sounds like this procedure would underestimate the variance in the number of mutations, since you remove the effect of random generation time. Is this effect small?
\end{point}

\reply{
}

\begin{point}{}
    Can the authors provide a bit more intuition behind the patterns of variation seen in generation time, census population size, and variance in the number of offspring with respect to neighborhood size seen in Figure 2? For example, it is not obvious to me why the census population size, for example, should decline systematically with respect to neighborhood size. Presumably this isn't just due to the local demographic stochasticity. Could the authors briefly interpret the observed patterns or cite appropriate literature?
\end{point}

\reply{
}

\begin{point}{}
    Fig. 7D: I am surprised by the extent to which the observed values of -log10(p) fall below the y=x line. Particularly in the lower right panel for large neighbourhood sizes. I would expect that to be close to panmictic - why are the P-values underdispersed? That seems like a potential bug, or else something weird is going on.
\end{point}

\reply{
}

\begin{point}{}
    Lines 706-716, It might be worth citing Haworth et al Nature Communications 2019 (https://doi.org/10.1038/s41467-018-08219-1) who do the proposed test (GWAS for birth location) in UK Biobank to illustrate the population structure.
\end{point}

\reply{
}

\begin{point}{}
    The analysis and discussion around the effect of GWAS is focused on PCA correction. Do mixed models help at all?
\end{point}

\reply{
}

\begin{point}{}
    The github link to the code didn't work for me. I assume it will be made public later, but at this point I can't tell whether the code is available/useable.
\end{point}

\reply{
}

%%%%%%%%%%%%%%%%%%%
\reviewersection{3}

\begin{quote}
    The present study deals with a "hot topic" in spatial population genetics. Most
    inferential and descriptive methods in statistical spatial population
    genetic rely on a discrete approximation of space and it is not clear what
    impact this approximation may have when individuals migrate along a
    continuum instead. Spatial patterns in sampling is also another major issue
    which is often simply dismissed, mainly because of the paucity of
    statistical methods to deal with it. This work touches on these important
    issues in a timely manner.
\end{quote}

\begin{quote}
    Although I was enthusiastic about the topic, I was quite disappointed with
    the core of the study, i.e., the forward-in-time simulation of populations in
    continuous space. The field has been struggling with this issue for decades --
    examples of spectacular failures like the Wright-Malecot model (see
    Felsenstein's "pain in the torus' article, 1975) or, more recently, the
    ``mugration' or ``discrete trait analysis' model in phylodynamics (see De Maio
    et al. 2015) have probably mostly harmed our research field -- that one cannot
    make the economy of using a sound probabilistic model for generating
    geo-referenced genetic data. It does not seem to be the case here
    unfortunately.  
\end{quote}

\begin{point}{}
    First, the simulation starts with individuals distributed uniformly at random in space. Is there any indication that the three-step algorithm used here maintains this distribution during the course of evolution? If it does not, then is there any stationary regime and how many generations does one need to wait before reaching it? I do appreciate that the competitive interaction term was introduced in order to avoid seeing the ``clumping' of individuals that hampers the Wright-Malecot model. Yet, just because there are no such clusters does not mean that the spatial distribution of individuals reaches a stable regime and that the distribution reached, if any, is reasonable from a biological perspective.
\end{point}

\reply{
}

Second, the demographic process used here involves birth and death of individuals. Does the population survive asymptotically or, like any birth-death process, eventually dies with probability one? In fact, one needs to know a little about the dynamics of the population size to decide whether the corresponding process is reasonable from a biological standpoint.
\begin{point}{}

\end{point}

\reply{
}

\begin{point}{}
    Third, it is not clear what the relationship between the expected lifespan and the probability of survival is. The expected lifespan, L, is first defined as the inverse of the expected number of offspring produced by a parent. The authors also define the probability of survival of a given individual at a given point in space, $p_i$. Hence, the expected lifespan at a point in space (and time) is the mean of a geometric distribution with parameter $p_i$, i.e., $1/p_i$. Now, it is far from being obvious what the relationship between these two approaches for defining the expected lifespan actually is.
\end{point}

\reply{
}

\begin{point}{}
    Also, the web page https://github.com/petrelharp/spaceness does not seem to exist so that I was not able to experiment with the forward-in-time generator used here unfortunately.
\end{point}

\reply{
}

\begin{point}{}
    All in all, more efforts need to be made here in my opinion to show that the forward-in-time simulations generate sensible outcomes. Sensible in terms of the behavior of the population demography at equilibrium (provided such equilibrium indeed exists) along with that of the spatial distribution of individuals. The authors could provide some guarantee of the good behavior of their model as evidenced from simulations using a broad range of parameter values for generating data. Alternatively, they could elect to use the spatial-Lambda-Fleming-Viot model for their simulations, which, in my opinion would seem the most sensible option given that (1) it is possible to run backward-in-time simulations under this model, thereby saving a lot of computation time and (2) it is a well-studied model with good mathematical and biological properties and (3) it is implemented in a publicly available software program (https://github.com/jeromekelleher/discsim)
\end{point}

\reply{
}

\begin{point}{}
    Figure 2: I do not understand why the neighborhood size varies to the same extent in the random mating model as it does for the spatial model. For the random mating model, I would have expected the neighborhood size to be equal to the census size since all individuals have the same probability of being a parent of any given offspring. From lines 166->171, it is clear that the spatial model would converge to the random mating model when the mean parent-offspring distance tends to infinity only if we were to ignore the impact of range edges. I am thus wondering whether the variation of neighborhood size one observes in Fig 2 for the random mating model is just a consequence of border effects. If that is the case, then the authors should state it clearly and try to justify it from a biological perspective.
\end{point}

\reply{
}

\begin{point}{}
    Line 729-731: "Many more species occur in a middle range of neighborhood sizes between 100 and 1000 - a range in which spatial processes play a minor role in our analyses [...]". Do the authors think that the spatial processes would still play a minor role when neighborhood sizes exceed 100-1000 if the habitat was larger than that taken in the present simulations? It would also probably be useful to mention that neighborhood sizes given in Table 1 should be compared with extreme caution since the size of the corresponding habitats vary across species. More generally, I suspect that the size of the habitat has a substantial impact on the vast majority of statistics examined in this study. Indeed, the mean parent-offspring distance, which is at the core of the definition of Wright's neighborhood size, is only small or large relative to the size of the habitat.
\end{point}

\reply{
}

\begin{point}{}
    Line 753-757: please add a reference to Guindon, Guo and Welch (2016). This study clearly shows that population density and dispersal parameters are identifiable and can indeed be estimated in practice under the spatial Lambda-Fleming-Viot model.
\end{point}

\reply{
}

%%%%%%%%%%%%%%%%%%%
\reviewersection{4}

\begin{quote}
    The manuscript by Battey et al explores the consequence of a well-known
    violation to population genetic models: the fact that populations are
    spatially structured and mate along a geographical cline, rather than
    randomly. This topic is important, particularly in light of recent working
    describing how spatially correlated genetic and environmental impacts can
    confound some population genetic insights, such as positive selection for
    height in Europe. The analyses and investigations presented here are
    thorough and sensible, and my comments are primarily intended to broaden
    accessibility for this interesting topic.
\end{quote}

\begin{point}{}
    Introduction. The discussion is very clear, articulating the three primary goals of the project: the impact of failing to model spatial population structure on 1) population genetic summary statistics, 2) inference on demographic history from population genetic data, and 3) impacts on GWAS summary statistics. I found the discussion a bit easier to follow than the introduction and would suggest streamlining and introducing the topic a bit more. Since the paper follows the flow described in the discussion, it might help orient readers by introducing these topics in the same order.
\end{point}

\reply{
}

\begin{point}{}
    I agree that most modern work describes structure as discrete populations connected by migration. However, some methods/studies have explicitly modeled spatial structure, e.g. especially in ecology or using methods like dadi (diffusion approximations). Highlighting some examples of previously identified structure not possible to infer without modeling geography would be helpful to contextualize this work.
\end{point}

\reply{
}

### 2019-08-14 ms/review-responses.tex

\begin{point}{}
    There is some reference to spatial models using grids (e.g. Rousset 1997). Some additional discussion contextualizing more recent methods like EEMS that also construct demes and model migration through divergence between neighboring demes would be helpful and interesting.
\end{point}

\reply{
}

\begin{point}{}
    Demographic modeling. Both approaches tested, stairwayplot and SMC++, are most sensitive to older demographic events, and consequently are very noisy and underestimate effect population sizes, especially in smaller neighborhood sizes. Models that consider haplotype structure are much better suited to this time period. It would be helpful to either 1) discuss the varying time sensitivities of different classes of demographic inference methods and how spatial patterns of genetic variation would influence these inferences, or 2) apply a method of this class (many options, e.g. DoRIS, IBDNe, Tracts, Globetrotter, etc) and show how it performs.
\end{point}

\reply{
}

\begin{point}{}
    GWAS mixed models. To what extent can spatial signals (e.g. corner, patchy) be corrected with mixed models, e.g. with PCs and PC-adjusted GRM as in Conomos et al, 2016 using PC-AiR and PC-Relate)? Is patchiness related to dispersal? I'm curious how this relates to the predictive ability of GWAS phenotypes with some spatial association that may or may not be associated with environmental effects.
\end{point}

\reply{
}

\begin{point}{}
    Code availability. This github link doesn't work, but is important to be able to evaluate for review: https://github.com/petrelharp/spaceness
\end{point}

\reply{
}

\begin{point}{}
    Definitions and interpretations. There are quite a large number of metrics discussed in Figure 3B, and it's a lot to take in. It might be helpful to have a table with a reminder of what the metric is, its interpretation, and how it is computed.
\end{point}

\reply{
}

\begin{point}{}
    Notation: "Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$ in both the x and y coordinates. Each offspring is produced with a mate selected randomly from those within distance $3\sigma$, with probability of choosing a neighbor at distance $x$ proportional to $\exp(-x^2/2\sigma^2)$." I think x may be overloaded here, or I'm confused. Clarify?
\end{point}

\reply{
}

\begin{point}{}
    When introducing the "spatial model" as opposed to this "random model," the more concrete illustration in Figure 1 hasn't yet been referenced, which makes it harder to follow. It would be helpful to introduce this figure with the model. Additionally, when Figure 1 is introduced, the order is from right to left (random, then point, then midpoint). It would be helpful to rearrange the figure to mirror what's in the text.
\end{point}

\reply{
}

\begin{point}{}
    Not sure I follow this example: "Concretely, an individual at position (x, y) in a 50 × 50 landscape has mean phenotype 100 + 2x/5."
\end{point}

\reply{
}

\begin{point}{}
    Minor typo (through vs though): "This occurs because, even through the "population density" (K) and "mean lifetime" (L) parameters..."
\end{point}

\reply{
}

\begin{point}{}
    Define NS abbreviation in Figure 5.
\end{point}

\reply{
}

### 2019-08-14 ms/spaceness.tex
review response stuff

@@

To avoid questionable assumptions, we here used forward-time, individual-based simulations. By scaling the probability of survival in each timestep to local population density we shift reproductive output towards low-density regions, which prevents populations from clustering. Such models have been used extensively in ecological modeling but rarely in population genetics, where to our knowledge previous implementations of continuous space models have focused on a small number of genetic loci, which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.

@@

To maintain a stable population, mortality increases with local population density.
To do this we say that individuals at distance $d$ have a competitive interaction with strength $g(d)$,
where $g$ is the Gaussian density with mean zero and standard deviation $\sigma$.
Then, the sum of all competitive interactions with individual $i$ is $n_i = \sum_j g(d_{ij})$,
where $d_{ij}$ is the distance between individuals $i$ and $j$ and the sum is over all neighbors within distance $3 \sigma$.
Since $g$ is a probability density, $n_i$ is an estimate of the number of nearby individuals per unit area.
Then, given a per-unit carrying capacity $K$, the probability of survival until the next time step for individual $i$ is
\begin{equation}
    p_i 
    =
    \min\left( 0.95, \frac{1}{1 + n_i / (K (1+L))} \right) .
\end{equation}
We chose this functional form so that the equilibrium population density per unit area is around $K$,
and the mean lifetime is around $L$.

@@

We ran 400 simulations for the spatial and random-mating models on a square landscape of width $W=50$ 
with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), 
average lifetime $L=4$, 
genome size = $10^{8}$, 
recombination rate = $10^{-9}$, 
and drawing $\sigma$ values from a uniform distribution between 0.2 and 4. 
To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM 
and later applied mutations to the tree sequence with msprime's \texttt{mutate} function \citep{Kelleher2016}. 
Because msprime applies mutations proportionally to elapsed time,
we divided the mutation rate of $10^{-8}$ mutations per site per generation
by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time.
(We verified that this procedure produced the correct number of mutations by comparing to a subset of simulations with SLiM-generated mutations, which are applied only at meiosis.)
Simulations were run for 1.6 million timesteps (approximately $30N$ generations).

@@

Our demographic model includes parameters for population density ($K$),
mean life span ($L$), and dispersal distance ($\sigma$).
However, nonlinearity of local demographic stochasticity causes
actual realized averages of these demographic quantitites to deviate from the specified values
in a way that depends on the neighborhood size.
Therefore, to properly compare to theoretical expectations,
we empirically calculated these demographic quantities in simulations. 
We recorded the census population size in all simulations. 
To estimate generation times, we stored ages of the parents of every new individual born across 200 timesteps, 
after a 100 generation burn-in, and took the mean. 
To estimate variance in offspring number, 
we tracked the number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period, subset the resulting table to include only the last timestep recorded for each individual, and calculated the variance in number of offspring across all individuals in timesteps 50-100.
All calculations were performed with information recorded in the tree sequence,
using pyslim \citetext{\url{https://github.com/tskit-dev/pyslim}}.

@@

In all simulations, the phenotype of each individual
is determined by adding independent Gaussian noise with mean zero and standard deviation 10
to a mean that may depend on spatial position.
We adjust the geographic pattern of mean phenotype to create spatially autocorrelated environmental influences on phenotype.
In the first simulation of \emph{nonspatial} environments, the mean did not change,
so that all individuals' phenotypes were drawn independently from a Gaussian distribution with mean 110 and standard deviation 10.
Next, to simulate \emph{clinal} environmental influences on phenotype, 
we increased the mean phenotype from 100 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
Concretely, an individual at position $(x,y)$ in a $50 \times 50$ landscape has mean phenotype $100 + 2x / 5$.
Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype 
for individuals with both $x$ and $y$ coordinates below 20 to 120 (two standard deviations above the rest of the map). 
Finally, in ``\emph{patchy}'' simulations we selected 10 random points on the map
and set the mean phenotype of all individuals within three map units of each of these points
to 120.

@@

We ran linear-model GWAS both with and without the first 10 principal components as covariates in PLINK and summarized results across simulations by counting the number of SNPs 
with $p$-value below $0.05$ after adjusting for an expected false positive rate of less than 5\% \citep{benjamini2001control}. 
We also examined $p$ values for systemic inflation by estimating the expected values from a uniform distribution (because no SNPs were used when generating phenotypes), plotting observed against expected values for all simulations, and summarizing across simulations by finding the mean $\sigma$ value in each region of quantile-quantile space. 
Results from all analyses were summarized and plotted with the ``ggplot2'' \citep{Wickham2016} and ``cowplot'' \citep{Wilke2019} packages in R \citep{Rcore2018}.

@@

Adjusting the spatial dispersal and interaction distance, $\sigma$,
has a surprisingly large effect on demographic quantities that are usually fixed in Wright-Fisher models --
the generation time, census population size, and variance in offspring number.
These are shown in Figure~\ref{fig:genparams}.
This occurs because,
even through the ``population density'' ($K$) and ``mean lifetime'' ($L$) parameters
were the same in all simulations,
the strength of stochastic effects depends strongly on $\sigma$.
For instance, 
the population density near to individual $i$ (denoted $n_i$ above)
is computed by averaging over roughly $N_W = 4 \pi K \sigma^2$ individuals,
and so has standard deviation proportional to $1/\sqrt{N_W}$ --
it is more variable at lower densities.
(Recall that $N_W$ is Wright's neighborhood size.)
Since the probability of survival is a nonlinear function of $n_i$,
actual equilibrium densities and lifetimes differ from $K$ and $L$.
This is the reason that we included \emph{random mating} simulations --
where mate choice and offspring dispersal are both nonspatial --
since this should preserve the random fluctuations in local population density
while destroying any spatial genetic structure.
We verified that random mating models retained no geographic signal
by showing that summary statistics did not differ significantly between sampling regimes
(Table~\ref{table:sampling}), unlike in spatial models (discussed below).

@@

\section{Acknowledgements}
We thank Brandon Cooper, Matt Hahn, Doc Edge, and others for reading and thinking about this manuscript. 
We thank the Hearth for having such good, nearby coffee, and
Falling Sky Brewing for creative support during manuscript drafting.
CJB and ADK were supported by NIH award R01GM117241.

@@

>>>>

%%%%%%%%%%%
% for responses to reviewers
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses
\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\input{review-response-commands}
\newcommand{\responsefile}{review-responses.tex}  % name of the review reponses file

@@

To avoid questionable assumptions, we here used forward-time, individual-based simulations. By scaling the probability of survival in each timestep to local population density we shift reproductive output towards low-density regions, which prevents populations from clustering. 
Such models have been used extensively in ecological modeling but rarely in population genetics, where to our knowledge previous implementations of continuous space models have focused on a small number of genetic loci, which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms.  \revpoint{AE}{2}
By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.

@@

To maintain a stable population, mortality increases with local population density.
To do this we say that individuals at distance $d$ have a competitive interaction with strength $g(d)$,
where $g$ is the Gaussian density with mean zero and standard deviation $\sigma$.
Then, the sum of all competitive interactions with individual $i$ is $n_i = \sum_j g(d_{ij})$,
where $d_{ij}$ is the distance between individuals $i$ and $j$ and the sum is over all neighbors within distance $3 \sigma$.
Since $g$ is a probability density, $n_i$ is an estimate of the number of nearby individuals per unit area.
Then, given a per-unit carrying capacity $K$, \llabel{ll:K_defn}
the probability of survival until the next time step for individual $i$ is
\begin{equation}
    p_i 
    =
    \min\left( 0.95, \frac{1}{1 + n_i / (K (1+L))} \right) .
\end{equation}
We chose this functional form so that the equilibrium population density per unit area is close to $K$, \llabel{ll:K_again}
and the mean lifetime is around $L$.

@@

We ran 400 simulations for the spatial and random-mating models on a square landscape of width $W=50$  \revpoint{AE}{3}
with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), 
average lifetime $L=4$, 
genome size = $10^{8}$, 
recombination rate = $10^{-9}$, 
and drawing $\sigma$ values from a uniform distribution between 0.2 and 4. 
To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM 
and later applied mutations to the tree sequence with msprime's \texttt{mutate} function \citep{Kelleher2016}. 
Because msprime applies mutations proportionally to elapsed time,
we divided the mutation rate of $10^{-8}$ mutations per site per generation
by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time.
(We verified that this procedure produced the correct number of mutations by comparing to a subset of simulations with SLiM-generated mutations, which are applied only at meiosis.)
Simulations were run for 1.6 million timesteps (approximately $30N$ generations).

@@

Our demographic model includes parameters that control population density ($K$), \revpoint{1}{2}
mean life span ($L$), and dispersal distance ($\sigma$).
However, nonlinearity of local demographic stochasticity causes
actual realized averages of these demographic quantitites to deviate from the specified values
in a way that depends on the neighborhood size.
Therefore, to properly compare to theoretical expectations,
we empirically calculated these demographic quantities in simulations. 
We recorded the census population size in all simulations. 
To estimate generation times, we stored ages of the parents of every new individual born across 200 timesteps, 
after a 100 generation burn-in, and took the mean. 
To estimate variance in offspring number, 
we tracked the number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period, subset the resulting table to include only the last timestep recorded for each individual, and calculated the variance in number of offspring across all individuals in timesteps 50-100.
All calculations were performed with information recorded in the tree sequence,
using pyslim \citetext{\url{https://github.com/tskit-dev/pyslim}}.

@@

In all simulations, the phenotype of each individual
is determined by adding independent Gaussian noise with mean zero and standard deviation 10 \revpoint{1}{5}
to a mean that may depend on spatial position.
We adjust the geographic pattern of mean phenotype to create spatially autocorrelated environmental influences on phenotype.
In the first simulation of \emph{nonspatial} environments, the mean did not change,
so that all individuals' phenotypes were drawn independently from a Gaussian distribution with mean 110 and standard deviation 10.
Next, to simulate \emph{clinal} environmental influences on phenotype, 
we increased the mean phenotype from 100 on the left edge of the range to 120 on the right edge (two phenotypic standard deviations).
Concretely, an individual at position $(x,y)$ in a $50 \times 50$ landscape has mean phenotype $100 + 2x / 5$.
Third, we simulated a more concentrated ``\emph{corner}'' environmental effect by setting the mean phenotype 
for individuals with both $x$ and $y$ coordinates below 20 to 120 (two standard deviations above the rest of the map). 
Finally, in ``\emph{patchy}'' simulations we selected 10 random points on the map
and set the mean phenotype of all individuals within three map units of each of these points
to 120.

@@

We ran linear-model GWAS both with and without the first 10 principal components as covariates in PLINK and summarized results across simulations by counting the number of SNPs 
with $p$-value below $0.05$ after adjusting for an expected false positive rate of less than 5\% \citep{benjamini2001control}. 
We also examined $p$ values for systematic inflation \revpoint{1}{6}
by estimating the expected values from a uniform distribution (because no SNPs were used when generating phenotypes), plotting observed against expected values for all simulations, and summarizing across simulations by finding the mean $\sigma$ value in each region of quantile-quantile space. 
Results from all analyses were summarized and plotted with the ``ggplot2'' \citep{Wickham2016} and ``cowplot'' \citep{Wilke2019} packages in R \citep{Rcore2018}.

@@

Adjusting the spatial dispersal and interaction distance, $\sigma$,
has a surprisingly large effect on demographic quantities that are usually fixed in Wright-Fisher models --
the generation time, census population size, and variance in offspring number.
These are shown in Figure~\ref{fig:genparams}.
This occurs because,
even through the parameters $K$ and $L$ that control population density and mean lifetime respectively \llabel{ll:K_more}
were the same in all simulations,
the strength of stochastic effects depends strongly on $\sigma$.
For instance, 
the population density near to individual $i$ (denoted $n_i$ above)
is computed by averaging over roughly $N_W = 4 \pi K \sigma^2$ individuals,
and so has standard deviation proportional to $1/\sqrt{N_W}$ --
it is more variable at lower densities.
(Recall that $N_W$ is Wright's neighborhood size.)
Since the probability of survival is a nonlinear function of $n_i$,
actual equilibrium densities and lifetimes differ from $K$ and $L$.
This is the reason that we included \emph{random mating} simulations --
where mate choice and offspring dispersal are both nonspatial --
since this should preserve the random fluctuations in local population density
while destroying any spatial genetic structure.
We verified that random mating models retained no geographic signal
by showing that summary statistics did not differ significantly between sampling regimes
(Table~\ref{table:sampling}), unlike in spatial models (discussed below).

@@

\section{Acknowledgements}
We thank Brandon Cooper, Matt Hahn, Doc Edge, and others for reading and thinking about this manuscript. 
% We thank the Hearth for having such good, nearby coffee, and
% Falling Sky Brewing for creative support during manuscript drafting.
CJB and ADK were supported by NIH award R01GM117241. 
\revpoint{AE}{4}

@@

\newpage

\includereviews
